Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-06-19 manuscript/manuscript.Rmd
Work on real-life datasets.

In the previous section, we simulated data to assess shift-sensitive power transformations and their robustness against outliers. In this section, we assess these aspects using real datasets.

## Datasets with outliers

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for the fuel consumption data in the Top Gear dataset. ",
  "This data contains three strong outliers for highly efficient vehicles. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.58$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust shift-sensitive transformations. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.36$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

@@

The Top Gear dataset contains data on 297 cars that appeared on the BBC television show *Top Gear*. Fuel consumption contains outliers. Figure \ref{fig:top-gear} shows that the outliers cause non-robust transformations to fail to transform the data to a centrally normal distribution (empirical central normality test $p = 0.01$). Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

### Housing dataset

```{r housing-data, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for the fuel consumption data in the Top Gear dataset. ",
  "This data contains three strong outliers for highly efficient vehicles. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2021-kq]. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.58$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust shift-sensitive transformations. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.36$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

.plot_sacramento_house(
  plot_theme = plot_theme
)
```

modeldata::Sacramento$price

Ames housing data set, year built

### Other data ???

@@

We investigated the use of robust invariant power transformation as part of the machine learning process, and did not find a major positive or negative effect. A possible explanation is that numeric features with large location shifts ($|\mu| > 1000.0$ were uncommon. Of the 4886 numeric features in the 232 datasets with numeric features, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.

>>>>

## Robustness against outliers

In the previous section, we simulated data to assess invariant power transformations and their robustness against outliers. In this section, we assess these aspects using real data.

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for the fuel consumption data in the Top Gear dataset. ",
  "The dataset contains three strong outliers for highly fuel-efficient vehicles.",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The empirical central normality test indicates that the original data do not deviate significantly from central normality ($p=0.76$).",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.55$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.03$, and $p=0.86$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear*. Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles. Figure \ref{fig:top-gear} shows that the outliers cause non-robust transformations to fail to transform the data to a centrally normal distribution (empirical central normality test $p = 0.01$ and $p=0.03$ for conventional and invariant transformations, respectively). Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

### Ischemic stroke dataset

```{r ischemic-stroke, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for the maximum arterial wall thickness in the ischemic stroke dataset [@Romano2022-gq]. ",
  "This data contains two strong outliers. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The empirical central normality test indicates that the original data do not deviate significantly from central normality ($p=0.56$).",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2021-kq]. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.61$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust shift-sensitive transformations. ",
  "For these, the empirical central normality test yields $p=0.02$, and $p=0.10$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

.plot_ischemic_stroke(
  plot_theme = plot_theme
)
```
The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
The angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Figure \ref{fig:ischemic-stroke} show that non-robust transformation fail to produce a centrally normal distribution (empirical central normality test $p = 0.01$ and $p=0.02$ for conventional and invariant transformations, respectively). Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

## Invariance

Datasets that where invariance has better properties in terms of lambda and resulting mean and standard deviation.
modeldata::ischemic_stroke$age
modeldata::ames$Longitudinal

@@

We investigated the use of robust invariant power transformation as part of the machine learning process, and did not find a major positive or negative effect. A possible explanation is that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon. Of the 4886 numeric features in the 232 datasets with numeric features, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.
